Computing Vowel Elision in Yoruba Collocations

### 2023-05-07 paper.tex
Finish analysis, write conclusion (rough)

@@

\subsection{Approach}

We base our procedure, crucially, on bounded depth-first search. We first
describe the basic algorithm, before proceeding onwards to an implementation.
Viewed as a single function which takes an input and returns an output, our
analysis takes as input an underlying form and a set of ranked constraints,
and searches for the correct output form. The search for the correct output
is of course the crucial portion here, and is the part on which most of the
discussion will be spent.

@@

We now describe key details of our implementation. The full source code may be
found at \href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
Starting with the basics, Figure \ref{fig:impl-deletion} shows both our
representation of Yoruba words and our simple handling of deletion.
Specifically, a word is represented as a list of segments; this is denoted by
\texttt{Vec<Segment>} on line 2.
At this point, it is reasonable to ask: why not represent words simply
as a list of characters? What's the difference between a character and a segment?
This reflects a complication in computers' handling of text beyond the
simple English characters covered by the ASCII character encoding. Unicode, an
improvement of ASCII that handles many more languages beyond those which use the
Latin alphabet, would consider the accent mark and letter \textipa{O} in \textipa{\'O},
to be separate characters and not parts of the same. We must sidestep this issue by
parsing words into \textit{graphemes}, which accounts for the intuitive notion
of what a character is: \textipa{\'O} is a single grapheme.

With that technical hiccup out of the way, our implementation of deletion is much
simpler. It is interesting to note that within the context of our analysis,
deletion turns out to be the most important of the operations to implement, because
winning candidates are always generated via deletion (due to high-ranking \dep{}
and \ident{} constraints). We implement deletion by picking a random element of
the list of segments and removing it. Because we cluster characters by Unicode
graphemes, the probability of any single character being deleted is evenly
distributed between all in a given word.

@@

The other significant portion of our implementation, beyond generating
candidates, is evaluating them. This, of course, is done with constraints.
Figure \ref{fig:impl-onset} shows our implementation of the \onset{}
markedness constraint. On lines 2-7, we count the number of syllable
nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
\ref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
Onsets and nuclei are calculated, along with other linguistic data, upon
the creation of a \texttt{SyllabifiedCandidate}; we merely analyze them
here. The number of violations of \onset{} is given by the number of
nuclei minus the number of onsets. This will never be a negative number,
because the algorithm that marks onsets and nuclei will only create one
onset per nucleus, at most (because this is the maximum allowed by
Yoruba).

@@

Of course, no discussion of constraints would be complete without
\textit{faithfulness} constraints. These present an interesting challenge,
but we first provide an overview of the implementation, which is provided
in Figure \ref{fig:impl-dep}. On lines 5 and 6, we respectively turn the
original underlying form and the candidate surface form into their string
respresentations. We then use a diff algorithm on these strings on line 8,
which returns a sequence of operations which could be performed on the
original underlying form to morph it into the candidate surface form given
as input. On lines 10-13, we count the number of insertion operations
returned by the diff algorithm.

@@

In this section we will evaluate all of our collected Yoruba underlying
representations against the expected surface forms.

TODO: Do the evaluations now that the implementation is reasonably
complete!

@@

>>>>

% Maybe add some background as Yoruba

@@

\subsection{Approach}
\label{subsec:approach}

We base our procedure, crucially, on bounded depth-first search. We first
describe a basic and idealized algorithm, before proceeding onwards to an
implementation section where we will discuss certain shortcuts which can be
taken due to convenient properties of vowel elision in Yoruba. Viewed as a
single function which takes an input and returns an output, our analysis 
takes as input an underlying form and a set of ranked constraints, and
searches for the correct output form. The search for the correct output is
of course the crucial portion here, and is the part on which most of the
discussion will be spent.

@@

We now describe key details of our implementation. The full source code may be
found at \href{https://github.com/hemantgouni/linguistics-capstone}{github.com/hemantgouni/linguistics-capstone}.
Starting with the basics, Figure \pref{fig:impl-deletion} shows both our
representation of Yoruba words and our simple handling of deletion.
Specifically, a word is represented as a list of segments; this is denoted by
\texttt{Vec<Segment>} on line 2.
At this point, it is reasonable to ask: why not represent words simply
as a list of characters? What's the difference between a character and a segment?
This reflects a complication in computers' handling of text beyond the
simple English characters covered by the ASCII character encoding. Unicode, an
improvement of ASCII that handles many more languages beyond those which use the
Latin alphabet, would consider the accent mark and letter \textipa{O} in \textipa{\'O},
to be separate characters and not parts of the same. We must sidestep this issue by
parsing words into \textit{graphemes}, which accounts for the intuitive notion
of what a character is: \textipa{\'O} is a single grapheme.

With that technical hiccup out of the way, our implementation of deletion is much
simpler. It is interesting to note that within the context of our analysis,
deletion turns out to be the most important of the operations to implement, because
winning candidates are always generated via deletion (due to high-ranking \dep{}
and \ident{} constraints). In fact, in the context of our specific analysis, it is
easy to generate \textit{all possible} surface forms instead of limiting ourselves to
the previously mentioned 65536 output forms, simply by choosing to delete or not
delete each segment. This yields $2^n$ possible output forms for a given input
form, where $n$ is the number of segments in the input form. We implement deletion
by picking a random element of the list of segments and removing it. Because we
cluster characters by Unicode graphemes, the probability of any single character
being deleted is evenly distributed between all in a given word.

@@

The other significant portion of our implementation, beyond generating
candidates, is evaluating them. This, of course, is done with constraints.
Figure \pref{fig:impl-onset} shows our implementation of the \onset{}
markedness constraint. On lines 2-7, we count the number of syllable
nuclei in the input (a \texttt{SyllabifiedCandidate} from Figure
\pref{fig:impl-deletion}). On lines 9-13, we count the number of onsets.
Onsets and nuclei are calculated, along with other linguistic data, upon
the creation of a \texttt{SyllabifiedCandidate}; we merely analyze them
here. The number of violations of \onset{} is given by the number of
nuclei minus the number of onsets. This will never be a negative number,
because the algorithm that marks onsets and nuclei will only create one
onset per nucleus, at most (because this is the maximum allowed by
Yoruba).

@@

Of course, no discussion of constraints would be complete without
\textit{faithfulness} constraints. These present an interesting challenge,
but we first provide an overview of the implementation, which is provided
in Figure \pref{fig:impl-dep}. On lines 5 and 6, we respectively turn the
original underlying form and the candidate surface form into their string
respresentations. We then use a diff algorithm on these strings on line 8,
which returns a sequence of operations which could be performed on the
original underlying form to morph it into the candidate surface form given
as input. On lines 10-13, we count the number of insertion operations
returned by the diff algorithm.

@@

In this section we will evaluate all of our collected Yoruba underlying
representations against the expected surface forms. The constraints
used in our analysis, along with their rankings, are listed here:

\begin{enumerate}
    \item
        \begin{enumerate}
            \item \ident{}
            \item \dep{}
        \end{enumerate}
    \item
        \begin{enumerate}
            \item \onset{}
            \item \ssp{}
            \item \maxc{}
        \end{enumerate}
    \item
        \begin{enumerate}
            \item \maxplusv{}
        \end{enumerate}
    \item
        \begin{enumerate}
            \item \maxvplus{}
        \end{enumerate}
\end{enumerate}

The number under which each constraint is listed denotes its rank.
For example, \ident{} and \dep{} have the highest ranking, followed
by \onset{}, \ssp{}, and \maxc{}. This largely matches
Fig. \pref{tableau:summary-two}, but with the inclusion of an extra
\maxc{} constraint at rank two. This is necessary because,
otherwise, our program (correctly) finds deleting the input form
to be the optimal solution to the set of constraints given. That is,
neither of the rank 1 constraints (\ident{} and \dep{}), nor the
constraints immediately below them (\onset{} and \ssp{}), generate
violations for deleting words. On the other hand, a form with no
segments will \textit{always} generate less violations of \onset{}
and \ssp{} than one which does-- the mere presence of a vowel in a
form is sufficient to generate \ssp{} violations. Adding a \maxc{}
constraint ranked equally to \onset{} and \ssp{} addresses this.
Indeed, this is not an issue with our implementation, but a more
fundamental issue with our analysis which was found by it. In that
sense, we can see that the benefits of computationally encoding the
analysis are already emerging.

An additional point to note is that, when evaluating candidates,
instead of considering equally-ranked constraints individually, we
treat violations between them cumulatively. As a result, since
\ssp{} generates multiple violations at a time (for instance [i]
will generate 3 violations, per Figure
\pref{table:vowel-sonority-v2}), in order for \maxc{} to
adequately regulate against deletion, it must generate multiple
(specifically, 3) violations for each deletion. In addition,
\onset{}, being equally ranked to \ssp{} and \maxc{}, must also
do this. This quirk of our implementation is certainly
unconventional among the OT literature, and further study might
better support some of these choices (or provide more elegant
alternative solutions).

Beyond these issue, our analysis is sound. Given an underlying
Yoruba form, the program implementing it will produce the
appropriately vowel-elided surface form. In other words, given
the information in the `Underlying' columns of tables
\pref{fig:deletion-simple}, \pref{fig:deletion-except-1},
\pref{fig:deletion-except-1}, our program outputs the
form expected in the 'Surface' columns (excepting
\textipa{/n\'i-oko/} \rightarrow{} \textipa{[l\'oko]},
since the n \rightarrow{} l transition occurs for reasons
unrelated to our analysis; we also ignore tone markers
here, because it is a related but orthogonal phenomenon
unhandled by our original analysis).

In fact, it will even produce the intuitively expected surface
form when given underlying forms which do not exist. For
instance, given \textipa{[so-el\'u]} as an input, it will produce
\textipa{/sel\'u/}, due to \maxplusv{} regulating more
strictly against initial vowel deletion than deletion of
final vowels.

@@

In this paper, we presented a classic Optimality-Theoretic
analysis of conjoined vowel elision in Yoruba. We then
implemented it as a computer program, finding a subtle
issue in the analysis. With these issues addressed in the
implementation, our program outputs the correctly
vowel-elided surface form, given an underlying
representation. Future work could go in several interesting
directions. Within the domain of Yoruba, the original
analysis and its implementation could be extended to handle
the assimilatory strategy for dealing with sequences of
vowels at morpheme boundaries (as discussed in
\cite{ola2002yoruba}), as well as tone. This would likely
involve extending surface candidate generation to the full
idealized algorithm described in Section
\ref{subsec:approach}, contending with epenthesis,
metathesis, and segment changes in addition to deletion.
The implementation, as it exists, faithfully models the
parts of the algorithm in \ref{subsec:approach} that are
necessary in the context of Yoruba, so this would largely
be a case of extending the existing machinery.

\section{Appendix}
